Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-06-11 Response_to_Reviewers.tex

%\makelettertitle

%Dear Editor,

\justify

%Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''. The suggestions and comments will undoubtedly be of great value to the paper. It was updated accordingly.

@@

\medskip

\textbf{Response to reviewers}

\bigskip

%We now turn to point-by-point responses to referees comments.

\textbf{Point-by-point response to referees comments:}

@@

\medskip

\textbf{First referee:}

\medskip

%The paper describes an approach that takes spatial sensitivity analysis explicitly and thoroughly into account. Based on the criticism that most agent-based-modelling and simulation approaches do neglect a careful recognition of the spatial impact on social phenomena, it introduces a “spatial generator” whose task is to generate different possible spatial configurations that are then used as a comparative measurement in socio-spatial simulations to prove the influence of spatial structures. To evaluate the strengths and weaknesses of the use of the spatial generator tool the author(s) applies two classical approaches – Schelling’s segregation model and the Sugarscape model – as benchmark models.

\begin{enumerate}
 % The methodology described is very convincing as it allows the utilisation of a tool that takes the spatial conditions – as causes or effects or intermediates – actually explicitly into account. With this approach, the paper convincingly outlines a pragmatic application by creating typical spatial compositions in urban contexts that is monocentric, polycentric or discontinuous cities. From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator. For example, paragraph 2.8 mentions four dimensions of spatial structure but leave the reader alone with just the terms. Either describe these or omit it.
   \item \comment[JR]{will add some description on urban form indicators etc.}

  %This also applies to the three segregation indexes: what exactly do they measure?
  \item

  % The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added, for example, to delineate the terms “strength of attrac! tion” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity? If so, is it then in space and/or time?). 
  \item \comment[JR]{will add more precision in the description of the generator}

  %While the methodological part of the paper is convincing the theoretical and epistemological background is less so. In short, the paper would improve significantly, if most of the geographical and empirical associations would be left out. Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc. 
  \item \comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}

  % The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation. In so doing, the author(s) runs immediately into the problem of adequate definitions. In paragraph 1.2 the definition of geographical systems as “social agents interacting with one another, within a limited portion of space” is misleading. First, it is a tautology (geographical systems in space); secondly, geographical cannot be described with social agents, because geographical systems are composed of spatial units, their relations, and a spatially defined system border. Social systems, in turn, might be described with humans as elements, their relations, and again a socially defined system border. Moreover, with respect to sociological system theory (Luhmann 1993), space does not add any knowledge to the nature of social systems. This approach can be (and I would add: has to be) criticised, however, not by interming! ling fuzzily the social with the geographical, as happened with this definition. 
  \item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ? which is far from geographical in the sense he puts it indeed}

  % Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described. Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account. These interactions can then be associated with the material and relational dimension of geographical spaces. The question remains: what is the role of geographical systems? In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”. 
  \item

  % The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following. What is described here as “spatial system shape” is the geometry of spaces but not the geography of spaces. 
  % -> yes indeed
  \item

  % Even more critically is the arbitrary conflation of social and spatial units as happened in paragraph 2.1, as geographical entities are described with “people, housing, networks, etc.” One is inevitably asking oneself: do geographical entities comprise of simply everything? Obviously not, but the paper seems to argue this way. 
  \item

  % Due to this, attempts to explain the growth of existing cities should be avoided. The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless. The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive. Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate. Or, the results can be used as a starting point to dig deeper in geographical reasons, taking social, economic, cultural, and political, but also topographical issues into consideration. This geographical reasoning, however, is outside of the scope of this paper. 
  \item \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate}

   % Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.
   \item

\end{enumerate}

\bigskip
\bigskip

\textbf{Second referee:}

%This manuscript presents a methodology to explore the effects of spatial parameters in ABM. They basically explore what they called “generator parameters”, to distinguish them from other statistical ones, which include density spatial distribution and patterns. To do so, they compared phase diagrams and calculated the mean of final values. 
%As case-studies, they used the actual distributions of three European cities as a means to constraint the density generation in two archetypical model (Schelling's segregation model and Epstein and Axtell's Sugarscape).

\begin{enumerate}

  % While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations. The two models have different behavioural assumptions and represent a different 'interaction' context. How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes? 
  \item

  % In 4.1, the authors suggest potential developments and point to certain limitations. While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions. For instance, it would be useful to have suggestions on how to use this methodology to run empirical models. Suppose to have a nice dataset and a case of a urban context to be modelled. Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations? 
  \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}

  % Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), whcih could be interesting to see implemented in ABM. Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa. 
  \item

  % Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics. Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones! 
  \item

  % In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research. I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints. And perhaps, authors could suggest future line of research that could show-case their argument. 
  \item

%I appreciate that models are fully available to readers.

\end{enumerate}

%\makeletterclosing

### 2019-07-25 paper.tex
Update on Overleaf.

\begin{abstract}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in geographical models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change.
\end{abstract}

@@

Computer simulation has been recognised and is increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}. Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems. The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way space influences the outcomes of social processes modelled. We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependency \citep{arthur1994increasing}. Therefore, observing a system at different points in time does not equate to observing different systems at a single point in time. This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues about past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models. Ironically, these very models tend to be the models most frequently used in geosimulation.

Self-organization has been shown to be a central feature of geographical systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}. In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}. Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer. \\

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents tend to be approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of urban processes because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they might want to live in \citep{SpielmanHarrison2014}, or agents moving around will sense a different set of objects within the same fixed radius depending on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent initial space is therefore a central element of geographical simulation models. However, this step is rarely explicit. A meaningful way of representing initial space might be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour to different plausible spatial arrangements.

@@

In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape). In no way do we pretend to provide a full exploration of these two particular models, their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models by using a spatial generator to produce a variety of density grids, which are taken as input by simulation models at initialisation. The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Previous considerations on the effects of the spatial configuration in simulation models}

In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc. For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income. Dynamically, sprawl in American cities consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}. Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}. The discussion of these effects has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}. For example, \citet{Kwan2012} has argued for a careful examination of what she coins the 'uncertain geographic context problem' (UGCoP), i.e. of the spatial configuration of geographical units even when the size and delineation of the area are the same.\\

Considerations of such issues in the geosimulation literature are rather scarce. However, there have been some noticeable attempts at analysing the impact of three types of initial spatial characteristics on model outcomes:
\begin{itemize*}
\item The accuracy of geo-localised input data;
\item The shape, precision and boundaries of the modelled spatial system;
\item The degree of spatial heterogeneity modelled.
\end{itemize*}

\subsubsection{Geo-localised input data accuracy}

\citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and the scale of analysis. They provide a few examples on how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes. In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and the computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.

\subsubsection{Spatial system shape, precision, and boundaries}

\citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e. made equivalent to) to Axelrod Culture Model. \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favours segregationist behaviours. \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins. Similar issues were also dealt with in physical sciences: for example, \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance below a certain level. Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time. Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinations of tolerance threshold and vacancy density values.

\subsubsection{Spatial heterogeneity}

\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents). In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome. In a spatial epidemics model for an infectious disease, parameterised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.

Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) or on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time. More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.

@@

In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution aggregate model behaviours (e.g. the final segregation level for Schelling). We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space. In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
%\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
\includegraphics[width=0.9\textwidth]{figures/Fig1.png}
\caption{\textbf{General workflow.}} \label{fig:method}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Quantitative variation across grids}

@@

\subsection{Qualitative variations}
\label{sec:qualResults}

@@

We consider that the method presented in this paper holds great potential for strengthening geographical models' exploration. However, two limits and two areas of opportunities have still not been tackled.

>>>>

\begin{abstract}
Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which socio-spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change.
\end{abstract}

@@

Computer simulation has been recognised and is increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}. It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}. Simulation also appears as a way to overcome the difficult analytic resolution of many socio-spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems. The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way space influences the outcomes of social processes modelled. We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

Socio-spatial systems can be crudely described as social agents interacting with one another, within a limited portion of space. Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependency \citep{arthur1994increasing}. Therefore, observing a system at different points in time does not equate to observing different systems at a single point in time. This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}. Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues about past bifurcations, that can have had dramatic effects on the state of the system. Therefore, strong spatio-temporal path-dependencies in the trajectory of individual territories and changing social environments over time prohibits the use of ergodic models. Ironically, these very models tend to be the models most frequently used in geosimulation.

Self-organization has been shown to be a central feature of socio-spatial systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}. In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}. Complexity is partially due to bifurcations, which are determinant in socio-spatial systems~\citep{Wilson1981, Wilson2002}. Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer. \\

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most socio-spatial models, especially cellular automata, still represent cities as uniform grids of isotropic density. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents tend to be approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of urban processes because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they might want to live in \citep{SpielmanHarrison2014}, or agents moving around will sense a different set of objects within the same fixed radius depending on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent initial space is therefore a central element of socio-spatial simulation models. However, this step is rarely explicit. A meaningful way of representing initial space might be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour to different plausible spatial arrangements.

@@

%In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).

In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system. {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world. The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems. In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} among themselves and with their geographical environment. The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper. We present how to generate the geometric outputs to are used as initial spatial conditions in the model. To this end, we use the example of a density grid generator and feed its outputs to two application models (Schelling and Sugarscape). In no way do we pretend to provide a full exploration of these two particular models, of their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models generated systematically. The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of geometric structures (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{The effects of the space in simulation models}

\subsubsection{Spatial processes}
In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc. For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income. Dynamically, sprawl in American cities consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}. Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}. These effects of geographical space on social systems correspond to processes to be included in geosimulation models.\\

\subsubsection{Spatial representation}
A discussion of the effects of spatial encoding and representation has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}. For example, \citet{Kwan2012} has argued for a careful examination of what she coins the 'uncertain geographic context problem' (UGCoP), i.e. of the spatial configuration of geographical units even when the size and delineation of the area are the same. Considerations of such issues in the geosimulation literature are rather scarce. However, there have been some noticeable attempts at analysing the impact of three types of initial spatial characteristics on model outcomes:
\begin{itemize*}
\item The accuracy of geo-localised input data. \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and the scale of analysis. They provide a few examples on how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes. In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and the computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
\item The shape, precision and boundaries of the modelled spatial system. \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e. made equivalent to) to Axelrod Culture Model. \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents. \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favours segregationist behaviours. \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins. Similar issues were also dealt with in physical sciences: for example, \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance below a certain level. Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time. Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinations of tolerance threshold and vacancy density values.
\item The degree of spatial heterogeneity modelled. \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice. They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns. \citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents. They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile. \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents). In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome. In a spatial epidemics model for an infectious disease, parameterised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.
\end{itemize*}

\subsubsection{Spatial structures}
We can distinguish a last category of spatial effects in geosimulation, which are the geometric constraints of the environment modelled at initialisation on the course and output of the simulation. Our original contribution is to tackle this type of spatial effects. In this paper, we present an operation framework which allows to systematically measure the impact of the initial geometric structures on the aggregate behaviour of simulation models. We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.

%Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) or on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time. More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.

@@

@@

The general workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the spatial distribution of socio-spatial entities (people, housing, networks, etc.). In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\centering
%\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
\includegraphics[width=0.9\textwidth]{figures/Fig1_v2.png}
\caption{\textbf{General workflow.}} \label{fig:method}
\end{figure} %
%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Sensitivity analysis}

@@

\subsection{Variation by type of structure}
\label{sec:qualResults}

@@

We consider that the method presented in this paper holds great potential for strengthening socio-spatial models' exploration. However, two limits and two areas of opportunities have still not been tackled.
